Simulating Ultra High-Dimensional Multivariate Data

### 2020-12-03 notebooks/cmstats2020_talk.rmd
2nd draft CMStats 2020 talk

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
subtitle: "Using the `bigsimr` R package"
author:
  - Alfred G. Schissler
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Anna K. Panorska
  - Juli Petereit
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr-60x60.png
bibliography: ../bigsimr.bib
abstract: "It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$"
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
set.seed(12212020)
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
```

@@

## Problem we're addressing

- In this era of Big Data, it is critical to realistically simulate data.
- This is computationally challenging when data are correlated and (ultra-) high dimensional.
- Many domains are producing these type of data.
- A prototypical example is RNA-sequencing data derived from high-throughput biomedical experiments.

## Our goal

- Our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.
- The simulation replicates $N$ can be large (~100,000).
- And the dimension $d$ can be very large (~20,000).
- We propose a scaleable NORmal To Anything approach to this task .
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`.

@@

Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

GGally::ggpairs( data = small,
                # lower = list( continuous = wrap(lowerfun)),
                upper = list(
                    continuous = wrap('cor', method = "spearman")
                ), 
                ) + theme_bw()
```

Our purposed methodology should possess the following properties  
(adapted from [@Nik13a]):

@@

In multivariate analysis, an analyst must select a metric to quantify dependency.
The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

\begin{equation}
(\#eq:pearson)
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
\end{equation}

@@

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. 
Define

\begin{equation}
(\#eq:spearman)
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
\end{equation}

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$. 
Spearman's $\rho_S$ has an appealing correspondence as the Pearson's correlation coefficient on *ranks* of the values, thereby captures nonlinear yet monotone relationships.

Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.
By concordance we mean that orderings have the same direction (e.g., if $X_i < X_j$, then $Y_i < Y_j$) and is determined by the ranks of the values, not the values themselves.

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.
As we will describe more fully in the [Algorithms](#algorithms) section, this property is essential to scaleable match rank-based correlations with speed (SP1) and accuracy (SP2).

## Correspondence among Pearson, Spearman, $\tau$ correlations

This is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
But for **bivariate normal vectors**, however, the correspondence is well-known:

\begin{equation}
(\#eq:convertKendall)
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's $\rho$ [@K58],

\begin{equation}
(\#eq:convertSpearman)
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
\end{equation}

## Gaussian copulas

There is a strong connection of our simulation strategy to Gaussian **copulas** (see @Nelsen2007 for a technical introduction).
A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals.
This definition provides a powerful, natural way characterize joint probability structure.
Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function 
$C(u_1, \ldots, u_d)$ so that

\[
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
\]

A Gaussian copula is the case where all marginal CDFs $F_i$ are the standard normal CDF, $\Phi$. 
This representation corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$.
But since the marginals are standardized to have unit variance, this ${\bf R_P}$ is a Pearson correlation matrix. 
If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

\begin{equation}
(\#eq:gauss)
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where $\Phi(\cdot)$ is the standard normal CDF.
Note that the copula $C_{{\bf R}}$ is the familar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions.
For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)$ with $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$, via ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$ povides $Y_i \sim F_i, \, \forall \, i$ .

@@

The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
Specifically, the NORTA algorithm follows like this:

@@

@@

There are a few R packages for multivariate simulation.

@@

Now we describe `bigsimr::rvec`, our algorithm to generate random vectors.
It mirrors the classical NORTA algorithm above with some modifications for rank-based dependency matching:

@@

The pre-processing (Step 1) takes advantage of the closed-form relationships between $\rho_S$ and $\tau$ with $\rho_P$ for bivariate normal random variables via Equations \@ref(eq:convertSpearman) or \@ref(eq:convertKendall), respectively (implemented as `bigsimr::cor_covert`).

## Algorithm notes

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

# 3 The `bigsimr` R package

## Installation and setup

- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
  performance algorithms.

## Basic use illustrated through a minimal example

```{r ch030-basic-example-setup, echo=TRUE, message=FALSE, cache=F}
library(bigsimr)
library(tidyverse)
library(patchwork)
```

For simplicity and to provide a minimal working example, we'll consider bivariate simulation of temperature, in degrees Fahrenheit, and ozone level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE}
df <- airquality %>%
  select(Temp, Ozone) %>%
  drop_na()
```

```{r ch030-aq-glimpse}
glimpse(df)
```

Figure \@ref(fig:ch030-aq-joint-dist) visualizes the bivariate relationship between Ozone and Temperature.
We aim to simulate random two-component vectors mimicking this structure. 
The margins are not normally distributed, particularly the ozone level exhibits a strong positive skew.

```{r ch030-aq-joint-dist, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Bivariate scatterplot of Ozone vs. Temperature with estimated marginal densities. The data are left skewed tails and appear to be correlation.", cache=F}

p0 <- ggplot(df, aes(Temp, Ozone)) +
  geom_point(size = 1) +
  theme(legend.position = "none") +
  labs(x = "Temperature")

pTemp <- ggplot(df, aes(Temp)) + 
  geom_density() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

pOzone <- ggplot(df, aes(Ozone)) + 
    geom_density() +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

pTemp + plot_spacer() + p0 + pOzone + 
  plot_layout(widths = c(3,1), heights = c(1, 3)) 
```

Next, we specify the marginal distributions and correlation coefficient (both its type and magnitude).
Here the analyst is free to be creative.
For this example, we will not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

## Specifying marginal distributions

```{r ch030-aq-temp-pars, echo=TRUE, eval=TRUE}
df %>% 
  select(Temp) %>% 
  summarise_all(.funs = c(mean = mean, sd = sd))
```

```{r ch030-aq-ozone-pars, echo=TRUE}
mle_mean <- function(x) mean(log(x))
mle_sd <- function(x) mean( (log(x) - mean(log(x)))^2 )

df %>% 
  select(Ozone) %>% 
  summarise_all(.funs = c(meanlog = mle_mean, sdlog = mle_sd))
```

```{r ch030-margins-alist, echo=TRUE}
margins <- alist(
    qnorm(mean = 77.871, sd = 9.4855),
    qlnorm(meanlog = 3.419, sdlog = 0.7426),
)
```

```{r ch030-margins-mlist, echo=TRUE}
margins <- mlist(
  qnorm(mean = mean(df$Temp), sd = sd(df$Temp)),
  qlnorm(meanlog = mle_mean(df$Ozone), sdlog = mle_sd(df$Ozone))
)
margins
```

## Specifying correlation

```{r ch030-aq-cor, echo=TRUE}
type <- 'spearman'
(rho <- cor_fast(df, method = "spearman"))
```

## Checking the theoretical correlation bounds

```{r ch030-cor-bounds, echo=TRUE}
cor_bounds(margins = margins, type = type)
```

Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
By comparison, the Pearson correlation $\rho_P$ is restricted to [-0.8648, 0.8727] for these margins (see `bigsimr::cor_bounds`output below).

```{r ch030-cor-bounds-pearson, echo=TRUE}
cor_bounds(margins = margins, type = 'pearson')
```

## Simulating random vectors

Finally, we arrive at the main function of `bigsimr`, `rvec`.
Let's now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, rho, margins, type)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

```{r ch030-plot-sim, fig.cap="Contour plot and marginal densities for the simulated bivariate distribution of Air Quality Temperatures and Ozone levels. The simulated points mimic the observed data with respect to both the marginal characteristics and bivariate association."}
p1 <- df_sim %>%
  ggplot(aes(Temp, Ozone)) +
  geom_density_2d_filled() +
  theme(legend.position = "none") +
  labs(x = "Simulated Temperature", y = "Simulated Ozone") +
  scale_y_continuous(limits = c(0, max(df$Ozone))) +
  scale_x_continuous(limits = range(df$Temp))

p1Temp <- ggplot(df_sim, aes(Temp)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank())

p1Ozone <- ggplot(df_sim, aes(Ozone)) + 
  geom_density(alpha = 0.5, fill = "lightseagreen") + 
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank()) +
  coord_flip()

p1Temp + plot_spacer() + p1 + p1Ozone + 
  plot_layout(widths = c(3,1), heights = c(1, 3))
```
---

# Monte carlo experiments

## Bivariate experiments

@@

Table: Identical margin, bivariate simulation configurations to evaluate correlation matching accuracy and efficiency.

@@

For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds (minus an adjustment factor of $\epsilon=0.01$ to handle numeric issues when the bound is specified exactly).

## Bivariate result table

```{r ch040-combineBiSims, cache=F}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

```{r ch040-biMAEtable, results='asis', cache=F}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    ## filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

knitr::kable(tabMSE,
             format = 'latex',
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             booktabs = TRUE,
             caption = 'Average abolute error in matching the target dependency across the entire range of possible correlations for each bivariate marginal.',
             linesep = c('', '',  '\\addlinespace')
)
```

## Bivariate results figure

Displayed are the aggregated bivariate simulation results.
```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.asp = 1.30, fig.weight = 4, fig.align='center', fig.cap = "Bivariate simulations match specified correlations. The horizontal axis plots the specified target correlations across the entire range of possible correlations for each bivariate margin. Normal margins are plotted in dark blue, gamma in medium blue, and negative binomial in light blue. As the number of simulated vectors $B$ increases from left to right, the variation in estimated correlations (vertical axis) decreases. The dashed line indicates equality between the specified and estimated correlations. Only the normal margins match the Pearson correlations (top row) exactly across the entire range of correlations. The two non-normal bivariate random vectors experience attenuation (downward bias) for extreme Pearson correlations (due to limitations in our algorithm) and more restriction (due to the Frechet bounds). The rank-based correlations (bottom two rows) are matched exactly for all margins and obtain the full range of possible dependencies."}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

@@

## Scale up to High dimensions using RNA-seq data

- We have RNA-sequencing data for N=1212 breast cancer patients (TCGA BRCA).  
- There are 20,501 genes that have count data quantifying the expression of the
  each gene's mRNA.  
- We simulate N random vectors to simulate a *fake dataset* with heterogeneous
  negative binomial margins and correlations estimated from real data.

```{r, echo=F, out.width='80%', fig.align='center', fig.cap='Computation times as d increases. We filter to the top 1, 5, 10, 15, 20, 25% expressing genes (in terms of median expression'}
knitr::include_graphics("../fig/cpu-gpu-times.png")
```

# 4 - Examples

## Simulating High-Dimensional RNA-seq data

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```
## Prepare to estimate the marginal parameters.

We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.
The functions below help achieve this estimation and specifying the inputs for use in `bigsimr::rvec`.

@@

## Estimate the marginal parameters

@@

Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`:

## Run `rvec`

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
colnames(sim_nbinom) <- names(brca)
```

@@

## Results for three example genes

Figure \@ref(fig:ch050-simDataFig) displays the simulated counts and pairwise relationships for our example genes in Table \@ref(tab:ch010-realDataTab).
Simulated counts roughly mimic the observed data but with a smoother appearance due to the assumed parameter form and with less extreme points then the observed data (c.f. Figure \@ref(fig:ch010-realDataFig).)

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.'}
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

## Results for all genes

Displayed are the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.

```{r ch050-figBRCA, echo=F, out.width = '80%', fig.align='center', fig.cap="Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicated equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes)."}
knitr::include_graphics('../fig/ch050-figBRCA.png')
```

## Simulation-based joint probability calculations

To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:

@@

So we estimate

@@

## Evaluation of correlation estimation efficiency

```{r ch010-myM, include=FALSE, cache = TRUE}
m <- 10
```

>>>>

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
author:
  - Alfred G. Schissler
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Anna K. Panorska
  - Juli Petereit
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr.png
bibliography: ../bigsimr.bib
abstract: "It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$"
---

```{r LoadLib, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(tidyverse)
set.seed(12212020)
## devtools::install_github("ggobi/GGally")
## for ggpairs
library(GGally)
## xaringan::inf_mr('cmstats2020_talk.rmd')
```

@@

## Big Multivariate Data

- It is critical to realistically simulate data when conducting Monte Carlo studies and methods.
- But measurements are often **correlated**, **non-normal**, and **high dimensional** in this era of big data.
- One example are *RNA-seq data* obtained through high-throughput biomedical experiments.
- 3 out of 20,501 mRNA measurements from 1212 breast cancer patients (TCGA BRCA):

```{r, echo=F, out.width='60%', fig.align='center'}
knitr::include_graphics("../fig/realDataFig.png")
```

- Our goal is to simulate $N$ random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with **correlated** components.

Our purposed methodology should possess the following properties  
(adapted from @Nik13a):

## Our contribution

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.

@@

- Simulating high-dimensional, non-normal, correlated data motivates this work --- in pursuit of modeling RNA-sequencing (RNA-seq) data [@Wang2009b; @Conesa2016b] derived from breast cancer patients. 
- The RNA-seq data-generating process involves counting how often a particular messenger RNA (mRNA) is expressed in a biological sample.
- RNA-seq platforms typically quantify the entire transcriptome in one experimental run, resulting in high-dimensional data.
- For human derived samples, this results in count data corresponding to over 20,000 genes (protein-coding genomic regions) or even over 77,000 isoforms when alternatively-spliced mRNA are counted.
- Importantly, due to inherent biological processes, gene expression data exhibits correlation --- co-expression --- across genes [@BE07; @Schissler2018].

@@

if ( !file.exists( '../fig/realDataFig.png' ) ) {
    p <- GGally::ggpairs( data = small,
                                        # lower = list( continuous = wrap(lowerfun)),
                         upper = list(
                             continuous = wrap('cor', method = "spearman")
                         )) + theme_bw()
    ggsave('../fig/realDataFig.png', plot = p, dpi = 300)
}
knitr::include_graphics("../fig/realDataFig.png")
```

@@

- In multivariate analysis, an analyst must select a metric to quantify dependency.
- The most widely-known is the Pearson (product-moment) correlation coefficient that describes the linear association between two random variables $X$ and $Y$, and, it is given by

$$
\rho_P(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}}.
$$

@@

## Rank-based measures of dependency

Two nonparametric, rank-based measures common in practice are **Spearman's correlation** (denoted $\rho_S$) and **Kendall's $\tau$**.

- Spearman's:

$$
\rho_S(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right],
$$

\noindent where $(X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y$ with $X_2$ and $Y_3$ are independent of one other and of $(X_1, Y_1)$.

-  Kendall $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations $(X_i, Y_i)$ and $(X_j, Y_j)$.

Both $\tau$ and $\rho_S$ are **invariant to under monotone transformations** of the underlying random variates.

## Correspondence among Pearson, Spearman, $\tau$

- There is no closed form, general correspondence among the rank-based measures and the Pearson correlation coefficient, as the marginal distributions $F_i$ are intrinsic in their calculation.
- But for **bivariate normal vectors**, however, the correspondence is well-known for Kendall's $\tau$:

$$
\rho_{P} = sin \left( \tau \times \frac{\pi}{2} \right), 
$$

- and similarly for Spearman's $\rho$ [@K58],

$$
\rho_P = 2 \times sin \left( \rho_S \times \frac{\pi}{6} \right).
$$

@@

- The well-known NORTA algorithm [@Cario1997] can be used simulate a random vector $\bf Y$ with variance-covariance matrix $\Sigma_{\bf Y}$.
- Specifically, the NORTA algorithm follows like this:

@@

- Exact Pearson solutions are possible but require computing $d \choose 2$ double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j$.
- But even when closed form solutions matching exactly can be impossible (*NORTA-defective* correlation matrices) and the probability of this increases with $d$ @GH02.

@@

Here are some other  R packages for multivariate simulation.

@@

@@

# 4 - Monte carlo experiments and examples

## 4A - Bivariate studies

@@

- Identical margin, bivariate simulation configurations to provide a low-dimensional evaluation:

@@

- For each of the unique 9 simulation configurations above, we estimate the correlation bounds and vary $\rho$ along a sequence of 100 points evenly placed within the bounds.

```{r ch040-bPlot, cache=F, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 6, fig.align='center'}
# https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
# New facet label names
repsLabs <- paste0("B=", c("1,000", "10,000", "100,000") )
names(repsLabs) <- c(1000, 10000, 100000)
typeLabs <- c( 'Pearson', 'Spearman', 'Kendall' )
names(typeLabs) <- c( 'Pearson', 'Spearman', 'Kendall' )

@@

## 4A - Average absolute error in bivariate studies

```{r ch040-combineBiSims, cache=F, echo=F}
## compare relative differences in MSE for each dependency, compared to Pearson
allDat <- NULL
dat <- readRDS("../results/biNorm_sims.rds") %>% select( margins, type, N, rho, rho_hat  ) 
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biGamma_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
dat <- readRDS("../results/biNB_sims.rds") %>% select( margins, type, N, rho, rho_hat  )
allDat <- rbind( allDat, dat )
allDat$margins <- factor(allDat$margins, levels = c( 'norm', 'gamma', 'nbinom'),
                         labels = c( 'MVN', 'MVG', 'MVNB' ) )
allDat$type <- factor(allDat$type, levels = c( 'pearson', 'spearman', 'kendall' ),
                      labels = c( 'Pearson', 'Spearman', 'Kendall' ) )
allDat$N <- factor(allDat$N, levels = c( 1000, 10000, 100000) )
```

```{r ch040-biMAEtable, cache=F, echo=F, warning=F, message=F}
## tabMSE  <- allDat %>%
##    group_by( N, margins ) %>%
##    summarize( mse = mean( ( rho - rho_hat )^2 ) )
tabMSE  <- allDat %>%
    filter( N == 100000) %>%
    group_by( N, type, margins ) %>%
##     summarize( mse = mean( ( rho - rho_hat )^2 ) ) %>%
    summarize( MAE = mean( abs( rho - rho_hat )  ) )
## normMSE <- as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ] ) 
## ( tabMSE$mse - normMSE ) / normMSE
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) )
## ( tabMSE$mse - as.numeric( tabMSE[ tabMSE$margins == 'norm', 'mse' ]) ) / tabMSE$mse

knitr::kable(tabMSE,
             col.names = c("No. of random vectors",
                           "Correlation type",
                           "Distribution",
                           "Mean abs. error"),
             caption = '',
)
```

## 4B - Simulating High-Dimensional RNA-seq data

@@

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( nb_Rho <- bigsimr::cor_fast( brca, method = corType ) )
```
## 4B - Estimate the marginal parameters

- We use method of moments (MoM) to estimate the marginal parameters for the multivariate negative binomial model.
- The marginal distributions are from the same probability family (NB) yet are heterogeneous in terms of the parameters probability and size $(p_i, n_i)$ for $i,\ldots,d$.

@@

## 4B - Estimate the marginal parameters

@@

- Notably, the marginal NB probabilities $\hat{p}_i's$ are small --- ranging in $[`r min(probs)` , `r max(probs)`]$.
- This gives rise to highly variable counts and, typically, less restriction on potential pairwise correlation pairs.
- Once the functions are defined/executed to complete marginal estimation, we specify targets and generate the desired random vectors using `rvec`.

## 4B - Simulate RNA-seq data via `bigsimr::rvec`

```{r ch050-runBRCA-echo, echo=TRUE, eval=FALSE, cache = FALSE}
## Set the number of random vectors
n <- 10000
## construct margins
nb_margins <- make_nbinom_alist(sizes, probs)
## run sims
sim_nbinom <- rvec(n, nb_Rho, nb_margins, type = corType,
                   ensure_PSD = TRUE, cores = cores) 
```

@@

## 4B -  Results for the three example genes

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='50%'}
knitr::include_graphics("../fig/realDataFig.png")
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

## 4B - Results for all genes

<!-- 
Displayed are the aggregated results of our simulation by comparing the specified target parameter (horizontal axes) with the corresponding quantities estimated from the simulated data (vertical axes).
The evaluation shows that the simulated counts approximately match the target parameters and exhibit the full range of estimated correlation from the data.
Utilizing 15 CPU threads in a MacBook Pro carrying a 2.4 GHz 8-Core Intel Core i9 processor, the simulation completed just over of 2 minutes.
-->

```{r ch050-figBRCA, echo=F, fig.align='center', out.width='60%'}
knitr::include_graphics('../fig/ch050-figBRCA.png')
```

## 4C - Simulation-based joint probability calculations

- To conduct statistical inference a critical task is to evaluate the joint probability mass (or density) function:

@@

- So we may estimate via:

@@

## 4D - Assessing estimation efficiency

-  To assess the error in our correlation estimation above:
